---
title: "A2"
author: "Annie Liu"
date: "2023-02-28"
output: html_document
---
### Dataset Introduction

In A1, I analyzed the dataset with the GEO ascension GSE75011, a dataset called Transcriptional profiling of TH2 cells identifies pathogenic features associated with asthma that was associated with the publication "Transcriptional Profilign of Th2 Cells Identifies Pathogenic Features Associated with Asthma" (Seumois et al. 2016). 

Asthma and allergic rhinitis are conditions affecting millions of people around the world, and these conditions are often comorbid with similar immunological hallmarks; however, not every person who has allergic rhinitis develop asthma in their lifetime. This study looks at the transcriptional profiling of Th2 cell enriched CD4 memory cells in patients with asthma, rhinitis, and healthy controls to attempt to gain genetic insights on these conditions. 

In A1, the aforementioned dataset was cleaned, filtered, and normalized using the edgeR protocol. The dataset contained gene names that were converted to HUGO symbols and then mapped to ensembl IDs. After data cleaning, 12973 genes were across all the samples were left with 12161 genes that had HUGO and ensmebl ids.  


## Load Libraries

``` {r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

if (!requireNamespace("GEOmetadb", quietly = TRUE))
    BiocManager::install("GEOmetadb")

if (!requireNamespace("edgeR", quietly = TRUE)) 
    BiocManager::install("edgeR")

if (!requireNamespace("DESeq2", quietly = TRUE)) 
    BiocManager::install("DESeq2")

if (!requireNamespace("biomaRt", quietly = TRUE)) 
    BiocManager::install("biomaRt")

if (!requireNamespace("dplyr", quietly = TRUE))
    install.packages("dplyr")

if (!requireNamespace("ComplexHeatmap", quietly = TRUE))
    BiocManager::install("ComplexHeatmap")

if (!requireNamespace("circlize", quietly = TRUE))
    install.packages("circlize")

if (!requireNamespace("ggplot2", quietly = TRUE))
    install.packages("ggplot2")

library("BiocManager")
library("GEOmetadb")
library("edgeR")
library("biomaRt")
library("DESeq2")
library("dplyr") 
library(ComplexHeatmap)
library(circlize)
library(knitr)
library("ggplot2")

```


## Data Visualization
```{r}
# Add normalized_count_data in
norm_data <- read.csv(file.path(getwd(), "data"))
norm_data = subset(norm_data, select= -X)
head(norm_data)

# Create numerical matrix that we can create a heatmap from

heatmap_matrix <- norm_data[, 4:ncol(norm_data)]
rownames(heatmap_matrix) <- norm_data$gene_hgnc
colnames(heatmap_matrix) <- colnames(norm_data[,4:ncol(heatmap_matrix)])

heatmap_matrix <- t(scale(t(heatmap_matrix)))

if(min(heatmap_matrix) == 0){
  heatmap_col = colorRamp2(c( 0, max(heatmap_matrix)), c("white", "red"))
} else {
  heatmap_col = colorRamp2(c(min(heatmap_matrix), 0, max(heatmap_matrix)), c("blue", "white", "red"))
}

current_heatmap <- Heatmap(as.matrix(heatmap_matrix),
                           show_row_dend = TRUE, show_column_dend= TRUE, col=heatmap_col, show_column_names = TRUE, show_row_names=FALSE, show_heatmap_legend=TRUE)

current_heatmap


# MDS plot 

cond <- data.frame(lapply(colnames(norm_data)[4:83], 
        FUN=function(x){unlist(strsplit(x, 
                        split = "_"))[c(1,2)]}))
colnames(cond) <- colnames(norm_data)[4:83]
rownames(cond) <- c("patient_id", "condition")
cond <- data.frame(t(cond))

expr_data <- as.matrix(norm_data[,4:83])
rownames(expr_data) <- norm_data$gene_hgnc
d = DGEList(counts=expr_data, group=cond$condition)
MDS_by_cond_edgeR <- plotMDS(d, labels=rownames(cond), col = c("darkgreen", "blue", "red")[factor(cond$condition)])

```

Going by what is present in the paper, we know that the sample groups are important in determining its value. Since we have two experimental groups here to one control, I am going to investigate each experimental group in relation to the other samples.

## Fitting model and computing differential expression
```{r}
# model design 
conditions <- cond$condition
model_d <- model.matrix(~0 + conditions) 

head(model_d)

d <- estimateDisp(d, model_d)

# fit model 

fit <- glmQLFit(d, model_d)

# compute differential expression 

compare_HCtoAR <- makeContrasts(HCtoAR="conditionsHC-conditionsAR", levels = model_d)
qlf_HCtoAR <- glmQLFTest(fit, contrast = compare_HCtoAR)
tt_HCtoAR <- topTags(qlf_HCtoAR, adjust.method = "BH", n=nrow(d))


# compare AS to all
contrast_AS <- makeContrasts(hctoall="conditionsAS-(conditionsHC + conditionsAS)/2", levels=model_d)
qlf_AS_to_all <- glmQLFTest(fit, contrast = contrast_HC)


# apply Benjamin-Hochberg 
tt_AS_to_all <- topTags(qlf_AS_to_all, adjust.method = "BH", n=nrow(d))
# 1934
length(which(tt_AS_to_all$table$PValue<0.05))
# 435
length(which(tt_AS_to_all$table$FDR<0.05))
output_AS_to_all <- merge(norm_data[,2:3], tt_AS_to_all, by.y=0, by.x=1, all.y=TRUE)
output_AS_to_all <- output_AS_to_all[order(output_AS_to_all$PValue),]
head(output_AS_to_all)

# compare AR to all
contrast_AR <- makeContrasts(hctoall="conditionsAR-(conditionsHC + conditionsAS)/2", levels=model_d)
qlf_AR_to_all <- glmQLFTest(fit, contrast = contrast_AR)

# apply Benjamin-Hochberg 
tt_AR_to_all <- topTags(qlf_AR_to_all, adjust.method = "BH", n=nrow(d))
# 1179
length(which(tt_AR_to_all$table$PValue<0.05))
# 17
length(which(tt_AR_to_all$table$FDR<0.05))
output_AR_to_all <- merge(norm_data[,2:3], tt_AR_to_all, by.y=0, by.x=1, all.y=TRUE)
output_AR_to_all <- output_AR_to_all[order(output_AR_to_all$PValue),]
head(output_AR_to_all)
```


## Volcano Plot of Differentially Expressed Genes
```{r}
# Code adapted from the following https://www.nathalievialaneix.eu/doc/html/solution-edgeR-rnaseq.html
# Plot code adapted from the following https://biocorecrg.github.io/CRG_RIntroduction/volcano-plots.html

# TODO: Colour and label the volcano plots 

# Volcano Plot of AS to all 
volcano_Data_AS_to_All <- cbind(tt_AS_to_all$table$logFC, -log10(tt_AS_to_all$table$FDR))
colnames(volcano_Data_AS_to_All) <- c("logFC", "negLogPval")
plot(volcano_Data_AS_to_All, pch=20,
     main="Volcano Plot of AS to all groups Differential Expression")


# Volcano Plot of AR to all
volcano_Data_AR_to_All <- cbind(tt_AR_to_all$table$logFC, -log10(tt_AR_to_all$table$FDR))
colnames(volcano_Data_AR_to_All) <- c("logFC", "negLogPval")
plot(volcano_Data_AR_to_All, pch=20,
     main="Volcano Plot of AR to all groups Differential Expression")
```
## Heatmap of Top Hits
```{r}

```

## Differential Gene Expression Summary
**Calculate p-values for each of the genes in your expression set. How many genes were significantly differentially expressed? What thresholds did you use and why?**

In our comparison of AS to all, there were 1934 genes with p-values less than 0.05. In our comparison of AR to all, there were 1179 genes with p-values less than 0.05.

**Multiple hypothesis testing - correct your p-values using a multiple hypothesis correction method. Which method did you use? And Why? How many genes passed correction?**

I used Benjamin-Hochberg as a multiple hypothesis correction method because this is what the original publishers of the dataset used. In our comparison of AS to all, 435 genes that passed correction. In our comparison of AR to all, 17 genes passed correction. 

**Show the amount of differentially expressed genes using an MA Plot or a Volcano plot. Highlight genes of interest.**
See the section "Volcano Plot of Differentially Expressed Genes"

**Visualize your top hits using a heatmap. Do you conditions cluster together? Explain why or why not.**




## Thresholded over-representation analysis
**Which method did you choose and why?**
**What annotation data did you use and why? What version of the annotation are you using?**
**How many genesets were returned with what thresholds?**
**Run the analysis using the up-regulated set of genes, and the down-regulated set of genes separately. How do these results compare to using the whole list (i.e all differentially expressed genes together vs. the up-regulated and down regulated differentially expressed genes separately)?**

## Interpretation
Do the over-representation results support conclusions or mechanism discussed in the original paper?
Can you find evidence, i.e. publications, to support some of the results that you see. How does this evidence support your results.

## References

Seumois, G., Zapardiel-Gonzalo, J., White, B., Singh, D., Schulten, V., Dillon, M., Hinz, D., Broide, D. H., Sette, A., Peters, B., & Vijayanand, P. (2016). Transcriptional Profiling of Th2 Cells Identifies Pathogenic Features Associated with Asthma. Journal of immunology (Baltimore, Md. : 1950), 197(2), 655–664. https://doi.org/10.4049/jimmunol.1600397

