---
title: "A1"
author: "Annie Liu"
date: "2023-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data set 
GSE Ascension: GSE138614
Data set title: Next-generation sequencing study in multiple sclerosis white matter brain lesions
Associated Publication: Molecular signature of different lesion types in the brain white matter of patients with progressive multiple sclerosis


### Step 1: Extracting the Dataset

Download the necessary packages needed for analysis.

``` {r}

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

if (!requireNamespace("GEOmetadb", quietly = TRUE))
    BiocManager::install("GEOmetadb")

if (!requireNamespace("edgeR", quietly = TRUE)) 
    BiocManager::install("edgeR")

library("BiocManager")
library("GEOmetadb")
library("edgeR")
```

Fetch the data set.

``` {r}
datadir <- "/home/rstudio/projects/GSE138614"

sfiles = getGEOSuppFiles('GSE138614')
fnames = rownames(sfiles)
# there is only one supplemental file
expressionData= read.delim(fnames[1],header=TRUE,
                check.names = FALSE)

```

### Step 2: Clean Dataset


``` {r}




```

### Step 3: Apply normalization to dataset

``` {r}




```

### Step 4: Map Identifiers 

``` {r}




```


### Step 5: Document 
What are the control and test conditions of the dataset?
Why is the dataset of interest to you?
Were there expression values that were not unique for specific genes? How did you handle these?
Were there expression values that could not be mapped to current HUGO symbols?
How many outliers were removed?
How did you handle replicates?
What is the final coverage of your dataset?

### References 


