---
title: "Assignment 1"
author: "Annie Liu"
date: "2023-02-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installing Packages Needed to Look for Data

We will be installing BiocManager and GEOmetadb in order to look for quality datasets. We will also be setting up GEOmetadb

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("GEOmetadb", quietly = TRUE))
    BiocManager::install("GEOmetadb")

if(!file.exists('GEOmetadb.sqlite')) GEOmetadb::getSQLiteFile()

library("GEOmetadb")

```

## Connect and Start Looking Through Data
Connecting to GEOmetadb and started exploration of GEOmetadb based on Week 3 Lectures
```{r}
library("GEOmetadb")
con <- dbConnect(SQLite(),'GEOmetadb.sqlite')
dbListFields(con, 'gpl')

uniq_tech <- dbGetQuery(con,'select distinct technology from gpl')
mod_table <- cbind(uniq_tech[1:(nrow(uniq_tech)/2),1],
       uniq_tech[((nrow(uniq_tech)/2)+1):nrow(uniq_tech),1])
knitr::kable( mod_table, format = "html")

num_uniq_tech_human <- dbGetQuery(con,'select technology,count(*) 
                                  as num_plat 
                                  from gpl 
                                  where organism = "Homo sapiens" 
                                  group by technology  
                                  order by num_plat desc')
colnames(num_uniq_tech_human)[2] <- "Num_Platforms"
dim(num_uniq_tech_human)


sql <- paste("SELECT DISTINCT gse.title,gse.gse, gpl.title",
             "FROM",
             "  gse JOIN gse_gpl ON gse_gpl.gse=gse.gse",
             "  JOIN gpl ON gse_gpl.gpl=gpl.gpl",
             "WHERE",
             "  gpl.title LIKE '%microarray%' ", #<<
             sep=" ")
rs <- dbGetQuery(con,sql)
dim(rs)
```

## Searching for a dataset

We want to find a dataset that has the following:
*with good coverage;
* not much older than ten years (quality!);
* with sufficient numbers of replicates;
* collected under interesting conditions;
* mapped to unique human gene identifiers

I am interested in high throughput data of asthma.
```{r}

sql <- paste("SELECT DISTINCT gse.title,gse.gse, gpl.title,",
             " gse.submission_date,",
             " gse.supplementary_file",
             "FROM",
             "  gse JOIN gse_gpl ON gse_gpl.gse=gse.gse",
             "  JOIN gpl ON gse_gpl.gpl=gpl.gpl",
             "WHERE",
             "  gse.submission_date > '2013-01-01' AND",
             "  gse.title LIKE '%brain%' AND", 
             "  gpl.organism LIKE '%Homo sapiens%' AND",
             "  gpl.technology LIKE '%high-throughput sequencing%' ",
             "  ORDER BY gse.submission_date DESC",sep=" ")

rs <- dbGetQuery(con,sql)
dim(rs)
knitr::kable(rs, format = "html")
unlist(lapply(rs$supplementary_file,
              FUN = function(x){x <- unlist(strsplit(x,";")) ;
              x <- x[grep(x,pattern="txt",ignore.case = TRUE)];
                tail(unlist(strsplit(x,"/")),n=1)})) 

series_of_interest <- rs$gse[grep(rs$supplementary_file,
                              pattern = "count|cnt",ignore.case = TRUE)]

shortened_filenames <- unlist(lapply(series_of_interest,
              FUN = function(x){x <- unlist(strsplit(x,";")) ;
              x <- x[grep(x,pattern= "count|cnt",ignore.case = TRUE)];
                tail(unlist(strsplit(x,"/")),n=1)}))
shortened_filenames[1:10]


num_series <- dbGetQuery(con, 
                         paste("select * from gsm where series_id in ('", 
                               paste(series_of_interest,collapse="','"), 
                               "')", collapse = ""))
gse.count <- as.data.frame(table(num_series$series_id))
series_of_interest_with_counts <- gse.count[which(gse.count$Freq>6),1]
gse.count[which(gse.count$Freq>6),]


```

# Retrieving the dataset 

We picked the dataset GSE138614
``` {r}
sfiles = getGEOSuppFiles('GSE138614')
fnames = rownames(sfiles)
# there is only one supplemental file
c = read.delim(fnames[1],header=TRUE,
                check.names = FALSE)
head(c)

```

